Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-04-15 pldi14/sec-performance.tex
Prepare integration: fix conflicts with polytable
Fewer undefined references

\begin{examples}
$\Derive{\MERGE} = \Lam {x\; \D x\; y \; \D y}{\Merge{\D x}{\D y}}$
is self-maintainable with the
change structure $\ChangeStruct{\Bag{S}}$ described in
\cref{ssec:change-structures}, because it does not use the base
inputs $x$ and $y$.
Other derivatives are self-maintainable only in certain contexts.
The derivative of element-wise function application
$\App*{\App\MAP f}{\mathit{xs}}$ ignores the original
value of the bag $\mathit{xs}$ if the changes to
%
$f$ are always nil, because the underlying primitive $\FOLDBAG$
is self-maintainable in this case (as discussed in next section).
%
We take advantage of this by implementing a specialized
derivative for $\FOLDBAG$.

We have seen in \cref{ssec:differentiation} that $\Derivative$
needlessly recomputes $\Merge\Xs\Ys$. However, the result is a
base input to $\FOLD'$.
%
In next section, we'll replace $\FOLD'$ by a self-maintainable
derivative (based again on $\FOLDBAG$) and will avoid this
recomputation.
\end{examples}

@@

\pg{As Klaus points out, this is not true until we fix the set of
  changes - discussed in \#265. So we should specify the set of
  changes, or use a different example. We should also motivate
  it.}
%
\pg{Agreed: Convert to bags and to our running example, using foldBag.}
To illustrate, suppose $\MAP$ is a primitive of sets of
integers,
and changes to sets consist of insertions and deletions only.
The term
\[
\Lam x {\Map {\Lam*n {n + 1}} x}
\]
contains a self-maintainable use of $\MAP$, because any change to
the input set~$x$, say $\{\App\INSERT5, \App\DELETE7\}$, can be
converted to an output change, say $\{\App\INSERT6,
\App\DELETE8\}$, without looking at $x$ itself. The use of $\MAP$
in the following term is not self-maintainable:
\[
\Lam x {\Map {\Lam*n {n + \Sum* x}} x}.
\]
Even one insertion to the input set~$x$ generates a sweeping
change over all elements of the output set that is impossible to
express in terms of insertions and deletions without knowledge of $x$.

>>>>

\begin{examples}
$\Derive{\MERGE} = \Lam {x\; \D x\; y \; \D y}{\Merge{\D x}{\D y}}$
is self-maintainable with the
change structure $\ChangeStruct{\Bag{S}}$ described in
\cref{ex:valid-bag-int}, because it does not use the base
inputs $x$ and $y$.
Other derivatives are self-maintainable only in certain contexts.
The derivative of element-wise function application
$\App*{\App\MAP f}{\mathit{xs}}$ ignores the original
value of the bag $\mathit{xs}$ if the changes to
%
$f$ are always nil, because the underlying primitive $\FOLDBAG$
is self-maintainable in this case (as discussed in next section).
%
We take advantage of this by implementing a specialized
derivative for $\FOLDBAG$.

\pg{broken reference}
We have seen in \cref{sec:differentiation-fold-example} that $\Derivative$
needlessly recomputes $\Merge\Xs\Ys$. However, the result is a
base input to $\FOLD'$.
%
In next section, we'll replace $\FOLD'$ by a self-maintainable
derivative (based again on $\FOLDBAG$) and will avoid this
recomputation.
\end{examples}

@@

\pg{As Klaus points out, this is not true until we fix the set of
  changes - discussed in \#265. So we should specify the set of
  changes, or use a different example. We should also motivate
  it.}
%
\pg{Agreed: Convert to bags and to our running example, using foldBag.}
To illustrate, suppose $\MAP$ is a primitive of sets of
integers,
and changes to sets consist of insertions and deletions only.
The term
\[
\Lam x {\MapT {\Lam*n {n + 1}} x}
\]
contains a self-maintainable use of $\MAP$, because any change to
the input set~$x$, say $\{\App\INSERT5, \App\DELETE7\}$, can be
converted to an output change, say $\{\App\INSERT6,
\App\DELETE8\}$, without looking at $x$ itself. The use of $\MAP$
in the following term is not self-maintainable:
\[
\Lam x {\MapT {\Lam*n {n + \Sum* x}} x}.
\]
Even one insertion to the input set~$x$ generates a sweeping
change over all elements of the output set that is impossible to
express in terms of insertions and deletions without knowledge of $x$.

### 2017-04-15 pldi14/sec-practice.tex
Fewer undefined references

As discussed in \cref{ssec:differentiation}, differentiation
supports abstraction, application and variables, but since
computation on base types is performed by primitives for those
types, efficient derivatives for primitives are essential for
good performance.

@@

\section{Case study}
\label{sec:plugins}

@@

\pg{Why does 'target language' show up all of a sudden? Shouldn't
  we say that we generate Scala code for execution
  \textbf{elsewhere}? Right now, we say it in next section. Confusing.}
%
\Cref{fig:primitives} shows a simplified
Scala implementation of the primitives
used in \cref{fig:case-study-pseudocode}.
As bag primitives, we provide constructors and a fold operation,
following \citet{GlucheGrust97Incr}. The constructors for bags
are $\Empty$ (constructing the empty bag), $\SINGLETON$
(constructing a bag with one element), $\MERGE$ (constructing the
merge of two bags) and $\NEGATE$ ($\Negate{b}$ constructs a bag
with the same elements as $b$ but negated multiplicities); all but $\SINGLETON$ represent
abelian group operations.
%
Unlike for usual ADT constructors, the same bag can be
constructed in different ways, which are equivalent by the equations defining abelian groups;
for instance, since $\MERGE$ is commutative, $\Merge{x}{y} = \Merge{y}{x}$.
%
Folding on a bag will represent the bag through constructors in
an arbitrary way, and then replace constructors with arguments;
to ensure a well-defined result, the arguments of fold should
respect the same equations, that is, they should form an abelian group;
for instance, the binary operator should be commutative.
%
Hence, the fold operator $\FOLDBAG$ can be defined to take a
function (corresponding to $\SINGLETON$) and an abelian group
(for the other constructors). $\FOLDBAG$ is then defined by equations:
%
\begin{alignat*}{2}
&  \HasType{\FOLDBAG}{\mathbf{Group}\; \Gt \to (\Gs \to \Gt) \to &&\Bag[\Gs] \to \Gt}\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Empty}
    &&= e \displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Merge* {b_1} {b_2}}
   &&= \App{\FoldBag{g}{f}}{b_1} \displaybreak[0]\\
&&& \boxplus
  \App{\FoldBag{g}{f}}{b_1} \displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Negate* b}
   && = \boxminus \; \App*{\FoldBag{g}{f}}{b}\displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\SingletonT*{v}}
   &&= \App{f}{v}
\end{alignat*}
%
If $g$ is a group,
these equations specify $\App{\FOLDBAG} g$ precisely~\citep{GlucheGrust97Incr}.
%
Moreover, the first three equations mean that $\FoldBag{g}{f}$ is the \emph{abelian group
  homomorphism} between the abelian group on bags and the group
$g$ (because those equations coincide with the definition).
%
\Cref{fig:primitives} shows an implementation of $\FOLDBAG$ as
specified above.
Moreover, all functions which deconstruct a bag can be expressed in
terms of $\FOLDBAG$ with suitable arguments.
%
For instance, we can sum the elements of a bag of integers with
$\FoldBag{\Term{gZ}}{\Lam*{x}{x}}$, where
$\Term{gZ}$ is the abelian group on integers defined in
\cref{ssec:change-structures}.
Users of $\FOLDBAG$ can define different abelian groups to specify
different operations (for instance, to multiply floating-point numbers).

If $g$ and $f$ do not change, $\FoldBag{g}{f}$ has a self-maintainable
derivative.
By the equations above,
\pg{Non-standard alignment, because the last line doesn't fit.}
\begin{align*}
& \FoldBag{g}{f} \Update*{b}{\D b}\displaybreak[0]\\
=\;& \FoldBag{g}{f} {(\Merge b {\D b})}\\
=\;& \App{\FoldBag{g}{f}}{b}  \boxplus \App{\FoldBag{g}{f}}{\D b} \\
=\;& \Update{\App{\FoldBag{g}{f}}{b}}{\Term{GroupChange} \; g \; \left(\App{\FoldBag{g}{f}}{\D b}\right)}
\end{align*}
We will describe the $\Term{GroupChange}$ change constructor in a moment.
Before that, we note that as a consequence, the derivative of $\FoldBag{g}{f} $ is
\[
\Lam{b \; db} \Term{GroupChange} \; g \; \left(\App{\FoldBag{g}{f}}{\D b}\right)\text{,}
\]
and we can see it does not use $b$: as desired, it is
\emph{self-maintainable}. Additional restrictions are require to
make $\FOLDMAP$'s derivative self-maintainable. Those restrictions require the
precondition on $\Term{mapReduce}$ in
\cref{fig:case-study-pseudocode}. $\Term{foldMapGen}$ has the
same implementation but without those restrictions; as a
consequence, its derivative is not self-maintainable, but it is more generally applicable.
Lack of space prevents us from giving more details.

To define $\Term{GroupChange}$, we need a suitable erased change
structure on $\Gt$, such that $\UPDATE$ will be equivalent to
$\boxplus$. Since there might be multiple groups on $\Gt$, we
\emph{allow the changes to specify a group}, and have
$\UPDATE$ delegate to $\boxplus$ (which we extract by pattern-matching on the group):
\begin{align*}
& \Change{\Gt} = \Term{Replace}\; \Gt \mid \Term{GroupChange} \Abelian*{\Gt} \Gt \\
& \Update{v}{(\Term{Replace} \; u)} = u\\
& \Update{v}{(\Term{GroupChange} \; (\boxplus, \Term{inverse}, \Term{zero})\; dv)} = v \boxplus dv\\
& \Diff{v}{u} = \Term{Replace}\; v
\end{align*}
That is, a change between two values is either simply the new
value (which replaces the old one, triggering recomputation),
or their difference (computed with abelian group
operations, like in the changes structures for groups from
\cref{ssec:change-structures}. The operator $\DIFF$
does not know which group to use, so it does not take advantage
of the group structure.
However, $\FOLDBAG$ is now able to generate a group change.
% Rewrite:
%derivatives of primitives like $\FOLDBAG$ can
%use the group structure they have available when producing changes.
%
\pg{Clarify about user-defined groups.}

We rewrite $\Program$ in terms of $\FOLDBAG$ to take advantage of
group-based changes.
{\DeriveProgramEnv
\begin{align*}
&\Term{id}=\Lam{x}{x}\\
&G_+ = (\mathbb Z, +, -, 0)\\
&\Program = \Lam{\Xs}{\Lam{\Ys}{\FOLDBAG~G_+~\Term{id}~(\Merge\Xs\Ys)}}\\
&\Derive\Program=\\
&\zero
\Lam{\Xs}{\Lam{\DXs}{}}\Lam{\Ys}{\Lam{\DYs}{}}\\
&\one
\FOLDBAG'~G_+~G_+'~\Term{id}~\Term{id}'\\
&\two
(\Merge\Xs\Ys)\\
&\two
(\MERGE'~\Xs~\DXs~\Ys~\DYs)
\end{align*}
}%
It is now possible to write down the derivative of $\FOLDBAG$.
{\DeriveProgramEnv
\begin{align*}
&\text{(if static analysis detects that $\D G$ and $\D f$
are nil changes)}\\
&\FOLDBAG'=\Derive{\FOLDBAG}=\\
&\zero
\Lam{G}{\Lam{\D G}{\Lam{f}{\Lam{\D f}{
\Lam{\Zs}{\Lam{\DZs}{}}}}}}\\
&\one
\Term{GroupChange}~G~
(\FOLDBAG~G~f~\DZs)
\end{align*}
}%
We know from \cref{ssec:plugin} that
\[
\MERGE'=\Lam{u}{\Lam{\D u}{\Lam{v}{\Lam{\D v}{\Merge{\D u}{\D v}}}}}.
\]
Inlining $\FOLDBAG'$ and $\MERGE'$ gives us a more readable term
$\beta$-equivalent to the derivative of $\Program$:
{\DeriveProgramEnv
\begin{align*}
&\Derive\Program=\\
&\zero
\Lam{\Xs}{\Lam{\DXs}{}}\Lam{\Ys}{\Lam{\DYs}{}}
%
\FOLDBAG~G_+~\Term{id}~
(\MERGE~\DXs~\DYs).
\end{align*}
}%

>>>>

As discussed in \cref{sec:informal-derive}, differentiation
supports abstraction, application and variables, but since
computation on base types is performed by primitives for those
types, efficient derivatives for primitives are essential for
good performance.

@@

\section{Case study}
\label{sec:plugins}
\pg{change naming for derivatives, they're dfold not fold' now.}

@@

\pg{Why does 'target language' show up all of a sudden? Shouldn't
  we say that we generate Scala code for execution
  \textbf{elsewhere}? Right now, we say it in next section. Confusing.}
%
\Cref{fig:primitives} shows a simplified
Scala implementation of the primitives
used in \cref{fig:case-study-pseudocode}.
As bag primitives, we provide constructors and a fold operation,
following \citet{GlucheGrust97Incr}. The constructors for bags
are $\Empty$ (constructing the empty bag), $\SINGLETON$
(constructing a bag with one element), $\MERGE$ (constructing the
merge of two bags) and $\NEGATE$ ($\Negate{b}$ constructs a bag
with the same elements as $b$ but negated multiplicities); all but $\SINGLETON$ represent
abelian group operations.
%
Unlike for usual ADT constructors, the same bag can be
constructed in different ways, which are equivalent by the equations defining abelian groups;
for instance, since $\MERGE$ is commutative, $\Merge{x}{y} = \Merge{y}{x}$.
%
Folding on a bag will represent the bag through constructors in
an arbitrary way, and then replace constructors with arguments;
to ensure a well-defined result, the arguments of fold should
respect the same equations, that is, they should form an abelian group;
for instance, the binary operator should be commutative.
%
Hence, the fold operator $\FOLDBAG$ can be defined to take a
function (corresponding to $\SINGLETON$) and an abelian group
(for the other constructors). $\FOLDBAG$ is then defined by equations:
%
\begin{alignat*}{2}
&  \HasType{\FOLDBAG}{\mathbf{Group}\; \Gt \to (\Gs \to \Gt) \to &&\Bag[\Gs] \to \Gt}\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Empty}
    &&= e \displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Merge* {b_1} {b_2}}
   &&= \App{\FoldBag{g}{f}}{b_1} \displaybreak[0]\\
&&& \boxplus
  \App{\FoldBag{g}{f}}{b_1} \displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\Negate* b}
   && = \boxminus \; \App*{\FoldBag{g}{f}}{b}\displaybreak[0]\\
&  \App{\FoldBag {g @ (\_, \boxplus, \boxminus, e)}{f}}{\SingletonT*{v}}
   &&= \App{f}{v}
\end{alignat*}
%
If $g$ is a group,
these equations specify $\App{\FOLDBAG} g$ precisely~\citep{GlucheGrust97Incr}.
%
Moreover, the first three equations mean that $\FoldBag{g}{f}$ is the \emph{abelian group
  homomorphism} between the abelian group on bags and the group
$g$ (because those equations coincide with the definition).
%
\Cref{fig:primitives} shows an implementation of $\FOLDBAG$ as
specified above.
Moreover, all functions which deconstruct a bag can be expressed in
terms of $\FOLDBAG$ with suitable arguments.
%
For instance, we can sum the elements of a bag of integers with
$\FoldBag{\Term{gZ}}{\Lam*{x}{x}}$, where
$\Term{gZ}$ is the abelian group induced on
integers by addition $(\mathbb{Z}, +, 0, -)$.
%
Users of $\FOLDBAG$ can define different abelian groups to specify
different operations (for instance, to multiply floating-point numbers).

If $g$ and $f$ do not change, $\FoldBag{g}{f}$ has a self-maintainable
derivative.
By the equations above,
\pg{Non-standard alignment, because the last line doesn't fit.}
\begin{align*}
& \FoldBag{g}{f} \Update*{b}{\D b}\displaybreak[0]\\
=\;& \FoldBag{g}{f} {(\Merge b {\D b})}\\
=\;& \App{\FoldBag{g}{f}}{b}  \boxplus \App{\FoldBag{g}{f}}{\D b} \\
=\;& \Update{\App{\FoldBag{g}{f}}{b}}{\Term{GroupChange} \; g \; \left(\App{\FoldBag{g}{f}}{\D b}\right)}
\end{align*}
We will describe the $\Term{GroupChange}$ change constructor in a moment.
Before that, we note that as a consequence, the derivative of $\FoldBag{g}{f} $ is
\[
\Lam{b \; db} \Term{GroupChange} \; g \; \left(\App{\FoldBag{g}{f}}{\D b}\right)\text{,}
\]
and we can see it does not use $b$: as desired, it is
\emph{self-maintainable}. Additional restrictions are require to
make $\FOLDMAP$'s derivative self-maintainable. Those restrictions require the
precondition on $\Term{mapReduce}$ in
\cref{fig:case-study-pseudocode}. $\Term{foldMapGen}$ has the
same implementation but without those restrictions; as a
consequence, its derivative is not self-maintainable, but it is more generally applicable.
\pg{No ``lack of space''}
Lack of space prevents us from giving more details.

To define $\Term{GroupChange}$, we need a suitable erased change
structure on $\Gt$, such that $\UPDATE$ will be equivalent to
$\boxplus$. Since there might be multiple groups on $\Gt$, we
\emph{allow the changes to specify a group}, and have
$\UPDATE$ delegate to $\boxplus$ (which we extract by pattern-matching on the group):
\begin{align*}
& \Change{\Gt} = \Term{Replace}\; \Gt \mid \Term{GroupChange} \Abelian*{\Gt} \Gt \\
& \Update{v}{(\Term{Replace} \; u)} = u\\
& \Update{v}{(\Term{GroupChange} \; (\boxplus, \Term{inverse}, \Term{zero})\; dv)} = v \boxplus dv\\
& \Diff{v}{u} = \Term{Replace}\; v
\end{align*}
That is, a change between two values is either simply the new
value (which replaces the old one, triggering recomputation),
or their difference (computed with abelian group
operations, like in the changes structures for groups from
\cref{sec:change-structure-groups}. The operator $\DIFF$
does not know which group to use, so it does not take advantage
of the group structure.
However, $\FOLDBAG$ is now able to generate a group change.
% Rewrite:
%derivatives of primitives like $\FOLDBAG$ can
%use the group structure they have available when producing changes.
%
\pg{Clarify about user-defined groups.}

We rewrite $\Program$ in terms of $\FOLDBAG$ to take advantage of
group-based changes.
{\DeriveProgramEnv
\begin{align*}
&\Term{id}=\Lam{x}{x}\\
&G_+ = (\mathbb Z, +, -, 0)\\
&\Program = \Lam{\Xs}{\Lam{\Ys}{\FOLDBAG~G_+~\Term{id}~(\Merge\Xs\Ys)}}\\
&\Derive\Program=\\
&\zero
\Lam{\Xs}{\Lam{\DXs}{}}\Lam{\Ys}{\Lam{\DYs}{}}\\
&\one
\FOLDBAG'~G_+~G_+'~\Term{id}~\Term{id}'\\
&\two
(\Merge\Xs\Ys)\\
&\two
(\MERGE'~\Xs~\DXs~\Ys~\DYs)
\end{align*}
}%
It is now possible to write down the derivative of $\FOLDBAG$.
{\DeriveProgramEnv
\begin{align*}
&\text{(if static analysis detects that $\D G$ and $\D f$
are nil changes)}\\
&\FOLDBAG'=\Derive{\FOLDBAG}=\\
&\zero
\Lam{G}{\Lam{\D G}{\Lam{f}{\Lam{\D f}{
\Lam{\Zs}{\Lam{\DZs}{}}}}}}\\
&\one
\Term{GroupChange}~G~
(\FOLDBAG~G~f~\DZs)
\end{align*}
}%
We know from \cref{sec:derive-example-merge} that
\[
\MERGE'=\Lam{u}{\Lam{\D u}{\Lam{v}{\Lam{\D v}{\Merge{\D u}{\D v}}}}}.
\]
Inlining $\FOLDBAG'$ and $\MERGE'$ gives us a more readable term
$\beta$-equivalent to the derivative of $\Program$:
{\DeriveProgramEnv
\begin{align*}
&\Derive\Program=\\
&\zero
\Lam{\Xs}{\Lam{\DXs}{}}\Lam{\Ys}{\Lam{\DYs}{}}
%
\FOLDBAG~G_+~\Term{id}~
(\MERGE~\DXs~\DYs).
\end{align*}
}%
